Tesseroids: forward modeling gravitational fields in spherical coordinates

### 2015-03-12 manuscript.tex
Added back proper quotation marks

Previous work by \citet{Ku1977} investigated the use of the GLQ in gravity forward modeling.
\citet{Ku1977} numerically integrated the vertical component of the
gravitational acceleration of right rectangular prisms.
The author suggested that the accuracy of the GLQ integration depends on
the ratio between distance to the computation point and the distance between
adjacent point masses.
Based on this, \citet{Ku1977} proposed an empirical criteria that the distance
between adjacent point masses should be less than the distance to the computation point.
\citet{Asgharzadeh2007} used  this criteria for the GLQ integration of
the gravity gradient tensor of tesseroids.
To our knowledge, the analysis if this ad hoc criteria of
\citet{Ku1977} works well for gravity gradient components or for tesseroids has never been done before.

There has also been no attempt to quantify the error committed in the GLQ
integration when applying the criteria of \citet{Ku1977}.

@@

\begin{figure}
    \centering
    \includegraphics[width=0.4\textwidth]{figs/vary-height-and-order}
    \caption{
        Example of the effect of varying
        the computation height
        and the number of point masses in the Gauss-Legendre Quadrature.
        Black circles represent the horizontal location of the point masses.
        a) $g_{xy}$ calculated at $400\ km$ height using GLQ order 2
        ($2 \times 2 \times 2 = 8$ point masses).
        b) At $150\ km$ height and GLQ order 2,
        the result resembles that of
        four point masses instead of a single tesseroid.
        This effect was shown by \citet{Ku1977}.
        c) At $150\ km$ but with a higher GLQ order of 30. In (c) the horizontal locations of the point masses were not shown. Notice that the result shown in (c) is similar to that expected for a single mass source.
    }
    \label{fig:glqerrorsample}
\end{figure}

@@

The advantage of this adaptive discretization is
that the number of point masses is only increased
in regions where the tesseroid are
closer to the computation point.
Notice that the alternative approach of
simply increasing the order of the GLQ
would increase the number of point masses
evenly throughout the whole tesseroids.

@@

\emph{Tesseroids} consists of command-line programs
written in the C programming language.
The package includes programs to calculate
the gravitational fields of tesseroids and
rectangular prisms (in both Cartesian and spherical coordinates),
as well as model and computation grid generation.
\todo[inline]{Valeria: Leo, I did not understand "as well as model and computation grid generation"}
\todo[inline]{Leo: I mean that there are programs to generate the model files and the grid files. This can be removed without any problem.}
All programs receive input through
command-line arguments and the standard input channel (STDIN)
and output the results through the standard output channel (STDOUT).
For example,
the command to generate a regular grid with $NLON \times NLAT$ points,
calculate on it $g_z$ and $g_{zz}$ caused by
the tesseroids in a file "MODELFILE'',
and save the results to a file called "OUTPUT''
is:

@@

Our definition of the dimensions of the tesseroid
(the "side lengths'' of \citet{Li2011})
along longitude, latitude, and radius, respectively, are
(Figure~\ref{fig:division}a)

@@

\noindent
for each $i \in (\lambda, \phi, r)$.
$D$ is a positive scalar
hereafter referred to as the "distance-size ratio''.
If the inequality holds for all three dimensions,
the tesseroid is not divided.
Thus, the distance-size ratio determines
how close the computation point can be
before we must divide the tesseroid.
The value of $D$ is indirectly responsible for
the accuracy of the solution and the computation time.
We will explore the relationship with the accuracy in the following section.

@@

This stack-based implementation
has some advantages over the original recursive implementation,
namely:
(1) It will be faster because it bypasses the overhead of function calls.
(2) It gives the developer more control over the recursion step.
In recursive implementations,
the developer has no control over
the maximum number of consecutive recursive calls
(i.e., the "recursion depth").
This limit may vary with programming language,
compiler, and operating system.
Overflowing the maximum recursion depth
may result in program crashes,
typically with cryptic or inexistent error messages.
In the stack-based implementation,
the developer has complete control.
Overflowing of the stack can be handled gracefully
with an error message
or even performing a suitable approximation of the result.

@@

Figure~\ref{fig:glqerrorsample} shows that the largest errors are spread over
on top of the tesseroid.
Thus, calculating the tesseroid fields at a single point might not
capture the point of largest error.
Instead, we calculate the effect of the tesseroid model on a regular grid
of $10 \times 10$ points at different geographic locations
(Table~\ref{tbl:experiment}).
\todo[inline]{ VAL: A grid of 10 X  10  point covers a are of ??? X ???? km}
\todo[inline]{LEO: The grid location varied in each experiment of the numerical analysis. Table 1 shows the locations of the grids.
This is discussed below.
I'll add this information to the text.}
Fortunately, the symmetry of the shell allows us to consider the computation
point at any geocentric coordinate.
Therefore, the effect of the shell will be same along the entire grid.
We compute the differences between the effects of the shell and the tesseroid
model on the grid.
However, we will consider only the largest error in our analysis.
\todo[inline]{ VAL: I rewrite the above sentence by:
"We compute the differences between the effects of the shell and the tesseroid model on the grid; however, we consider  the largest error only.  }
\todo[inline]{LEO: Rewritten. Is it better now?}
We placed the grid on top of a particular tesseroid
to increase the chances of capturing the true largest integration error.
\todo[inline]{ VAL:I rewrite the above sentence as:
"We decided for the nonconservative procedure of placing the grid on top of a particular tesseroid to increase the chances
of capturing the true largest integration error"}
\todo[inline]{LEO: Changed. I don't think this is really
``nonconservative'' so I left it out.}

>>>>

Previous work by \citet{Ku1977} investigated the use of the GLQ in gravity forward modeling.
\citet{Ku1977} numerically integrated the vertical component of the
gravitational acceleration of right rectangular prisms.
The author suggested that the accuracy of the GLQ integration depends on
the ratio between distance to the computation point and the distance between
adjacent point masses.
Based on this, \citet{Ku1977} proposed an empirical criteria that the distance
between adjacent point masses should be less than the distance to the computation point.
\citet{Asgharzadeh2007} used  this criteria for the GLQ integration of
the gravity gradient tensor of tesseroids.
To our knowledge, an analysis of how well this ad hoc criteria of
\citet{Ku1977} works for gravity gradient components or for tesseroids
has never been done before.
There has also been no attempt to quantify the error committed in the GLQ
integration when applying the criteria of \citet{Ku1977}.
\todo[inline]{LEO: Valeria, I've made some modifications to your correction.}

@@

\begin{figure}
    \centering
    \includegraphics[width=0.4\textwidth]{figs/vary-height-and-order}
    \caption{
        Example of the effect of varying
        the computation height
        and the number of point masses in the Gauss-Legendre Quadrature.
        Black circles represent the horizontal location of the point masses.
        a) $g_{xy}$ calculated at $400\ km$ height using GLQ order 2
        ($2 \times 2 \times 2 = 8$ point masses).
        b) At $150\ km$ height and GLQ order 2,
        the result resembles that of
        four point masses instead of a single tesseroid.
        This effect was shown by \citet{Ku1977}.
        c) At $150\ km$ but with a higher GLQ order of 30.
        In (c) the horizontal locations of the point masses were not shown.
        Notice that the results shown in (c) are similar to that expected
        for a single mass source.
    }
    \label{fig:glqerrorsample}
\end{figure}

@@

The advantage of this adaptive discretization is
that the number of point masses is only increased
in parts of the tesseroid that are
closer to the computation point.
Notice that the alternative approach of
simply increasing the order of the GLQ
would increase the number of point masses
evenly throughout the whole tesseroid.
\todo[inline]{LEO: Valeria, this correction you made is wrong.
It is not ``in regions where the tesseroids are closer''.
It is actually ``in parts of the tesseroid that are closer''.
This is all happening inside a single tesseroid.
I have corrected the text above.}

@@

\emph{Tesseroids} consists of command-line programs
written in the C programming language.
The package includes programs to calculate
the gravitational fields of tesseroids and
rectangular prisms (in both Cartesian and spherical coordinates).
\todo[inline]{Valeria: Leo, I did not understand "as well as model and computation grid generation"}
\todo[inline]{Leo: I mean that there are programs to generate the model files and the grid files. This can be removed without any problem.}
All programs receive input through
command-line arguments and the standard input channel (``STDIN'')
and output the results through the standard output channel (``STDOUT'').
For example,
the command to generate a regular grid with $NLON \times NLAT$ points,
calculate on it $g_z$ and $g_{zz}$ caused by
the tesseroids in a file ``MODELFILE'',
and save the results to a file called ``OUTPUT''
is:

@@

Our definition of the dimensions of the tesseroid
(the ``side lengths'' of \citet{Li2011})
along longitude, latitude, and radius, respectively, are
(Figure~\ref{fig:division}a)

@@

\noindent
for each $i \in (\lambda, \phi, r)$.
$D$ is a positive scalar
hereafter referred to as the ``distance-size ratio''.
If the inequality holds for all three dimensions,
the tesseroid is not divided.
Thus, the distance-size ratio determines
how close the computation point can be
before we must divide the tesseroid.
The value of $D$ is indirectly responsible for
the accuracy of the solution and the computation time.
We will explore the relationship with the accuracy in the following section.

@@

This stack-based implementation
has some advantages over the original recursive implementation,
namely:
(1) It will be faster because it bypasses the overhead of function calls.
(2) It gives the developer more control over the recursion step.
In recursive implementations,
the developer has no control over
the maximum number of consecutive recursive calls
(i.e., the ``recursion depth'').
This limit may vary with programming language,
compiler, and operating system.
Overflowing the maximum recursion depth
may result in program crashes,
typically with cryptic or inexistent error messages.
In the stack-based implementation,
the developer has complete control.
Overflowing of the stack can be handled gracefully
with an error message
or even performing a suitable approximation of the result.

@@

Figure~\ref{fig:glqerrorsample} shows that the largest errors are spread over
on top of the tesseroid.
Thus, calculating the tesseroid fields at a single point might not
capture the point of largest error.
Instead, we calculate the effect of the tesseroid model on a regular grid
of $10 \times 10$ points at different geographic locations
(see Table~\ref{tbl:experiment}).
\todo[inline]{ VAL: A grid of 10 X  10  point covers a are of ??? X ???? km}
\todo[inline]{LEO: The grid location varied in each experiment of the numerical analysis. Table 1 shows the locations of the grids.
This is discussed below.
I'll add this information to the text.}
Fortunately, the symmetry of the shell allows us to consider the computation
point at any geocentric coordinate.
Therefore, the effect of the shell will be same along the entire grid.
We compute the differences between the effects of the shell and the tesseroid
model on the grid.
However, we will consider only the largest error in our analysis.
\todo[inline]{ VAL: I rewrite the above sentence by:
"We compute the differences between the effects of the shell and the tesseroid model on the grid; however, we consider  the largest error only.  }
\todo[inline]{LEO: Rewritten. Is it better now?}
We placed the grid on top of a particular tesseroid
to increase the chances of capturing the true largest integration error.
\todo[inline]{ VAL:I rewrite the above sentence as:
"We decided for the nonconservative procedure of placing the grid on top of a particular tesseroid to increase the chances
of capturing the true largest integration error"}
\todo[inline]{LEO: Changed. I don't think this is really
``nonconservative'' so I left it out.}
